Using Incremental Many-to-One Queries to Build a Fast and Tight Heuristic for A* in Road Networks

### 2021-11-16 paper/ch_potentials.tex
grammar checking

\begin{abstract}
We study exact, efficient and practical algorithms for route planning applications in large road networks.
Routing applications often require integrating the current traffic situation, planning ahead with traffic predictions for the future, respecting forbidden turns, and many other features depending on the exact application.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many applications.
A* is a classical approach to accelerate Dijkstra's algorithm.
A* can support many extended scenarios without much additional implementation complexity.
However, A*'s performance depends on the availability of a good heuristic that estimates distances.
Computing tight distance estimates is a challenge on its own.
On road networks, shortest paths can also be quickly computed using hierarchical speedup techniques.
They achieve speed and exactness but sacrifice A*'s flexibility.
Extending them to certain practical applications can be hard.
In this paper, we present Lazy RPHAST, an algorithm to efficiently compute many distance estimates towards a single target and use it as a A* heuristic.
As it uses Contraction Hierarchies (CH), a hierarchical technique, we call our A* heuristic CH-Potentials.
Our approach allows decoupling the supported applications from the hierarchical speed-up technique.
Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks and present a new pruning criterion for symmetrical bidirectional A*.
An extensive experimental study confirms the practicality of our approach for many applications.
\end{abstract}

@@

Unfortunately, for many real world applications, this basic graph model is too simplistic.
For realistic routing, many additional features need to be considered.
This includes turn costs and restrictions, live traffic, user preferences, and traffic predictions. %, temporary driving bans for certain types of vehicles.
Some applications may have additional application-specific requirements.
Extending Dijkstra's algorithm to support these features is usually easy.
Extending hierarchical speed-up techniques is often possible, too.
However, the algorithm development is vastly more complex.
For every feature, dedicated research papers exist that extend CH and other approaches.
Supporting the combination of several features is even harder.
For example, we are not aware of any work combining all features mentioned above.
In this article, we describe an algorithmic building block, that allows handling the combination of all above mentioned features -- and probably more.

@@

There is a lot of work that extends hierarchical speed-up techniques to more complex settings~\cite{bdgmpsww-rptn-16}.
For example, in~\cite{gv-errnt-11} turn information is integrated into CH.
The proposed approach achieves fast queries but preprocessing becomes an order of magnitude slower compared to classical CH.

@@

In the this section, we recall the central aspects of Dijkstra's classical shortest path algorithm~\cite{d-ntpcg-59} and introduce notation used throughout this paper.
Dijkstra's algorithm maintains an array of tentative distances $D$ and a priority queue $Q$ of nodes ordered by increasing distance from $s$.
Initially, all distances are set to $\infty$.
The start node distance $D[s]$ is set to zero and the queue is initialized with $s$.
In each iteration, the closest remaining node $u$ is popped from the queue and \emph{settled}.
Outgoing edges $uv$ are relaxed, i.e. $D[v]$ is set to $\min(D[v], D[u] + w(uv))$.
If the tentative distance at the head node $D[v]$ was improved, it's added to the queue or its position in the queue is adjusted.
Once $t$ was settled, $D[t]$ contains the shortest path distance between $s$ and $t$.
The search can stop when $t$ is settled.

@@

In a CH, each node gets a unique rank representing its ``importance''.
By convention, important nodes are usually depicted at the ``top'' while unimportant nodes are positioned at the ``bottom''.
The intuition is that dead-ends are unimportant while highways or bridges are important.
An edge goes \emph{up} when its tail node has lower rank than its head node.
\emph{Down} edges are defined analogously.
An \emph{up-down path} is a path where only one node $m$ is more important than both its neighbors.
$m$ is called the \emph{mid} node.
An \emph{up path} is a path where the last node is the mid node.
Similarly, the first node is the mid node of a \emph{down path}.
% Every up and down path is an up-down path.
%
In the preprocessing phase, first, a node ordering is computed heuristically.
Then, \emph{shortcut} edges are added to the input graph $G$ to obtain $G^+$.
This is done by repeatedly contracting the least important node and adding shortcuts between its neighbors.
%See~\cite{gssv-erlrn-12} for the details.
After the preprocessing, for every pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path in $G^+$ with the same length as a shortest path in $G$.
See Figure~\ref{fig:ch} for a proof sketch.
From every shortest path (red) in $G$, an up-down path of equal length in $G^+$ (blue) can be constructed.
Thus, we can restrict our search to up-down paths in $G^+$.
The search is bidirectional.
The forward search starts from $s$ and only follows up-edges.
Similarly, the backward search starts at $t$ and only follows down-edges in reversed direction.
The two searches meet at the mid node.
Pseudo-code for the backward search, i.e., the path from $m$ to $t$, is presented in Algorithm~\ref{algo:ch-backward}.
The forward search works analogously.
%
A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
On road networks, this is the case~\cite{gssv-erlrn-12,dgpw-crprn-13}.
On graphs with low treewidth, this is also the case~\cite{dsw-cch-15,hs-gbpo-18}.

@@

Some hierarchical techniques were extended to support a three phase setup.
The phases consist of a slow preprocessing phase, a faster \emph{customization} phase, and fast queries.
The preprocessing phase typically takes hours, the customization seconds, and the query milliseconds.
The preprocessing phase is independent of the weights and only exploits the graph topology.
Weights are introduces in the customization phase.
Paths are computed in the query phase.
This setup is useful to allow for weight changes that are common to all vehicles such as life traffic updates.
The first technique to follow this setup is CRP~\cite{dgpw-crprn-13}

Customizable Contraction Hierarchies~\cite{dsw-cch-15} (CCH) are a variant of Contraction Hierarchies that uses this three phase setup.
The key idea is to use a node importance order that is independent of the weight function.
A CCH is also a CH.
The final result of the preprocessing is also an augmented graph $G^+$.
It fulfills all necessary properties for CH query algorithms.
The CH query algorithms can therefore be applied without modifications.
However, the CCH preprocessing grants some additional useful properties.

@@

PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node (or vice versa, the reverse case works analogously).
This is sometimes denoted as a \emph{all-to-one} (or \emph{one-to-all}) problem.
The preprocessing phase remains the same as for CH.
The query phase is split into two steps.
The first step is analogue to the CH query:
From $t$, all reachable nodes via reversed down-edges are explored.
Algorithm~\ref{algo:ch-backward} shows this first step.
For the second step, PHAST organizes nodes into \emph{levels}.
These levels correspond to the importance ordering but allow multiple nodes in the same level.
No edge must connect two nodes within the same level.
PHAST then iterates over all CH levels from top to bottom.
In each iteration, all up-edges starting within the current level are followed in reverse.
After all levels are processed, the shortest distances from all nodes to $t$ were computed.
Pseudo-code is provided in Algorithm~\ref{algo:phast}.
PHAST is faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
It can also easily be parallelized.
% However, we will not consider parallelization in this paper.
We refer to~\cite{dgnw-phast-13} for an in-depth experimental performance analysis.
% Using PHAST, we can also compute a tight A* heuristic.
% In the query phase, we first run PHAST to compute the distances from every node to $t$ with respect to $w_\ell$ and store the result in an array $H$.
% Next, we run $A^*$ and implement the heuristic as a lookup in the array $H$.

@@

RPHAST~\cite{delling_et_al:OASIcs:2011:3266}, short for Restricted PHAST, is a PHAST extension for efficiently computing distances from a smaller set of source nodes to one target node (again, the reversed case works analogously), solving the \emph{many-to-one} problem.
Given a set of source nodes $S$, the first step is to copy all up-edges reachable only through up-edges from the source nodes into a new graph.
This step is called \emph{selection}.
Then, in the query step, a target $t$ is given and the PHAST algorithm is applied but the downward sweep (second step of the PHAST algorithm) is performed only on the restricted subgraph.
RPHAST is particularly effective when many targets are queried for the same source set $S$.
If the source set changes often, selection times may become problematic.
RPHAST can also be applied to the \emph{many-to-many} problem, computing a distance matrix between sets $S$ and $T$ of nodes.
Implemented with SIMD instructions it is, to the best of our knowledge, the fastest currently known algorithm for this problem setting.

@@

In this section, we discuss another very specific problem setting which naturally arises from A* heuristics.
Here, the target node is known in the selection step but the source nodes are only queried one after another.
We denote this problem as the \emph{incremental many-to-one} problem.
The first step is the \emph{target selection} where the target node is provided.
Then, an arbitrary number of source nodes $s_i$ are given one after another.
For each source the distance to the target has to be computed before the next source will be provided.

@@

Analogous to A*, our algorithm stores for every node $v$ a tentative distance $D[v]$.
Additionally, it maintains a minimum priority queue.
Diverging from A*, not all nodes can be pushed but every node has a tentative distance.

@@

Our algorithm differs from A* when removing a node $u$ from the queue.
A* iterates over the outgoing edges $(uv)$ of $u$ and tries to reduce $D[v]$ by relaxing $(uv)$.
If A* succeeds, $v$'s weight in the queue is set to $D[v]+h_t(v)$.
Our algorithm, however, behaves differently, if $d(v)\le 2$.
Our algorithm determines the longest degree two chain of nodes $u, v_1,\ldots, v_k, w$ such that $d(v_i)=2$ and $d(w) > 2$.
If our algorithm succeeds in reducing $D[v_1]$, it does not push $v_1$ into the queue.
Instead, it iteratively tries to reduce all $D[v_i]$.
If it does not reach $w$, then only $D$ is modified but no queue action is performed.
If $D[w]$ is modified and $d(w)>2$, $w$'s weight in the queue is set to $D[w]+h_t(w)$.

@@

To avoid the downsides of the average potential, we revisit symmetric bidirectional A* and propose a new pruning criterion.
We describe the idea for the forward search.
The pruning rule for the backward search is analogous.
The central idea consists of using information from the backward search to prune edges in the forward search.
We do not use the average heuristic.
Instead of a strong stopping criteria, we use a pruning rule that gets stronger the longer the search runs.
Such a pruning rule will eventually prune all remaining branches and stop the search.
The stopping criteria for each direction remain the same (unidirectional) as before.
However, usually the search stops early because the queues are empty.

@@

Unless stated differently, for all bidirectional A* variants, we always alternate between removing a node from the forward and the backward queues.
In our experiments, we also evaluate expanding the search with the smaller minimum queue key.
This sounds sensible but our experiments show in Table~\ref{tab:bidir_switching} that it is never beneficial in terms of running time.

@@

% Ich bleibe bei v_i hier, damit keine Verwechselung mit den freien Variablen x,y,z gibt.
A path with nodes $v_1, v_2,\ldots v_k$ has the following \emph{turn-aware weight}: \[
w_\ell(v_1 v_2) + \sum_{i=2}^{k-1}  w_t(v_{i-1},v_i,v_{i+1}) + w_\ell(v_i v_{i+1})
\]
The objective is to find a path between two given edges with minimum turn-aware weight.
The first term $w_\ell(v_1, v_2)$ is the same for all paths, as it only depends on the source edge.
It can thus be ignored during optimization.

@@

If all travel time functions fulfill the \emph{FIFO property}, this problem can be solved using a straightforward extension of Dijkstra's algorithm \cite{d-aassp-69}.
The necessary modification to A* is analogous.
Without the FIFO property the problem becomes NP-hard \cite{or-tnp-89}.
% The FIFO property states that it is never beneficial to wait at a node before entering an edge.
The FIFO property states that it is not possible to arrive earlier by departing later.
Formally stated, the following must hold $\forall e\in E,\tau\in \mathbb{R},\delta\in \mathbb{R}^+: w_q(e, \tau) \le w_q(e, \tau+\delta) + \delta$.
Our implementation stores edge travel times using piece-wise linear functions.
The A* search uses the tentative distance $\tau$ at a node $u$ when to evaluating the travel time of outgoing edges $u v$.
This strategy is very similar to TD-ALT~\cite{ndls-bastd-12,dw-lbrdg-07}.

@@

This setup provides an alternative way of integrating live traffic compared to the approach described in Section~\ref{sec:live-predicted-traffic}.
Changes to the live traffic situation are handled by changing $w_\ell$ via a customization.
Additional features such as turn restrictions can be implemented by adjusting $w_q$.
For severe traffic events, the three phase approach will be significantly faster than the one described in Section~\ref{sec:live-predicted-traffic}.
However, it comes at the expense of a more complex algorithm.
Further, integrating predicted traffic becomes difficult, as $w_q$ may not drop below the $w_\ell$ with integrated congestion.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{fig/lazy_rphast_inc.pdf}
\caption{
Average running times of incremental Lazy RPHAST while querying $|S| = 2^{14}$ from a ball of varying size $|B|$ on OSM Germany excluding selection times.
The upper figure contains the total elapsed running time.
The lower figure contains the averaged running time per source, i.e. $y/x$ from the upper figure.
Note the different y-axis scales and units.
}\label{fig:lazy_rphast_inc}
\Description{
Two line graphs with each eight lines for different ball sizes and algorithms.
The x-axis in both graphs is the number of nodes that were queried up to that point in the incremental query.
In each graph, all lines have a similar shape and appear as scaled variants of each other.
The bigger the ball size the slower the whole algorithm.
The CCH based variant is somewhat slower for the most part.
Only for small ball sizes it is marginally faster when computing distances for more then 4096 sources.
When the ball is the entire graph, the CH based variant takes around 2.5ms for the first 512 sources, 5ms for 2048 sources and 25ms for 16384 sources.
Conversely, the average time per query is around 100mus for the first few queries, then drops to around 5mus after 512 sources, down to 2mus after 16384 sources.
For the smallest ball size of 16384, everything is 1.5 orders of magnitude faster.
}
\end{figure}

@@

Table~\ref{tab:building_blocks} depicts the performance of unidirectional A* with different heuristics and optimizations.
We compare (C)CH-Potentials to three other heuristics.
The first heuristic is, the \emph{Zero} heuristic where $h(v)$ is $0$ for all nodes $v$.
This corresponds to using Dijkstra's algorithm.
Secondly, we compare against our own implementation of ALT~\cite{gw-cppsp-05}.
We use 16 landmarks generated with the avoid strategy~\cite{gw-cppsp-05} and activate all during every query.
Finally, we compare against a hypothetical \emph{Oracle-A*} heuristic.
This heuristic has instant access to a shortest distance array with respect to $w_\ell$, i.e. it is faster than the fastest heuristic possible in our model.
We fill this array before each query using a reverse Dijkstra search from the target node.
In this experiment, we do not account for the running time of Dijkstra's algorithm.
Thus, the reported running times of Oracle-A* do \emph{not} account for any heuristic evaluation.
(C)CH-Potentials compute the same distance estimates but the heuristic evaluation has some overhead.
Comparing against Oracle-A* allows us to measure this overhead.
No other heuristic, which only has access to the preprocessing weights, can be faster than Oracle-A*.

@@

In Table~\ref{tab:bidir} we investigate the effectiveness of bidirectional search compared to unidirectional search depending on the query weights.
Interestingly, only the zero heuristic and ALT consistently achieve speed-ups through bidirectional search on all query weights.
With $w_q = w_{\ell}$, unidirectional CH-Potentials are already optimal and only traverse the shortest path.
Here, bidirectional search will only introduce unnecessary overhead.
When query weights are scaled up uniformly, bidirectional search achieves some search space reduction but it is not enough to significantly reduce running times due to the overhead of running a second search.
This changes drastically when the query weight increases are applied non-uniformly in the third scenario.
Here only weights with a speed less than 80\,kph were scaled up.
This touches only the beginning and end of most shortest paths between randomly chosen nodes.
The middle part of the shortest paths will typically use faster edges like highways.
In this case, bidirectional (C)CH-Potentials are a factor of two faster than the unidirectional variant.
The reason for this is that the search space of a unidirectional search expands greatly while exploring the end of the path to the target where the reduced weights are bad.
In contrast, the bidirectional searches meet in the middle of the shortest path where the reduced weights are close to zero and thus avoid this expansion.
This is also the reason why ALT behaves like this for all query weights.
By construction, the ALT heuristic has better reduced weights for edges which are on many shortest paths like highways.
Conversely, unimportant edges have bad reduced weights.
This makes bidirectional search so critical for ALTs performance.
In contrast, having a potential as tight as CH-Potentials makes bidirectional search in many scenarios unnecessary.
Bidirectional search for CH-Potentials only pays of when the reduced weights exhibit certain very specific patterns.

@@

In the other scenarios, the performance of CH-Potentials strongly depends on the quality of the heuristic.
We measure this quality using the length increase of $w_q$ compared to $w_\ell$.
Forbidding highways results in the largest length increase and in the smallest speedup.
The other extreme are turn restrictions.
They have only a small impact on the length increase.
The achieved speedups are therefore comparable to CH speedups.
Since turn costs an restrictions appear mostly in the beginning and end of shortest-paths and not in the middle on highways, utilizing bidirectional search results in even better speed-ups.
Mapbox live traffic has a length increase of around 15\%, which yields running times of 130\,ms.
Applying bidirectional CH-Potentials in this case is actually detrimental to the performance because the bad reduced edge weights appear in the middle of shortest paths.
The same is the case for forbidden tunnels or highways.
Applying CCH-Potentials customized to the current traffic (or the metric with forbidden edges) to these problems yields running times similar to the unmodified case and speed-ups of more than three orders of magnitude.
With bidirectional CCH-Potentials, one can easily additionally support turn costs and still have running times of a few milliseconds without any adjustments to the customization.

>>>>

\begin{abstract}
We study exact, efficient and practical algorithms for route planning applications in large road networks.
Routing applications often require integrating the current traffic situation, planning ahead with traffic predictions for the future, respecting forbidden turns, and many other features depending on the exact application.
While Dijkstra's algorithm can be used to solve these problems, it is too slow for many applications.
A* is a classical approach to accelerate Dijkstra's algorithm.
A* can support many extended scenarios without much additional implementation complexity.
However, A*'s performance depends on the availability of a good heuristic that estimates distances.
Computing tight distance estimates is a challenge on its own.
On road networks, shortest paths can also be quickly computed using hierarchical speedup techniques.
They achieve speed and exactness but sacrifice A*'s flexibility.
Extending them to certain practical applications can be hard.
In this paper, we present Lazy RPHAST, an algorithm to efficiently compute many distance estimates towards a single target and use it as an A* heuristic.
As it uses Contraction Hierarchies (CH), a hierarchical technique, we call our A* heuristic CH-Potentials.
Our approach allows decoupling the supported applications from the hierarchical speed-up technique.
Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks and present a new pruning criterion for symmetrical bidirectional A*.
An extensive experimental study confirms the practicality of our approach for many applications.
\end{abstract}

@@

Unfortunately, for many real world applications, this basic graph model is too simplistic.
For realistic routing, many additional features need to be considered.
This includes turn costs and restrictions, live traffic, user preferences, and traffic predictions. %, temporary driving bans for certain types of vehicles.
Some applications may have additional application-specific requirements.
Extending Dijkstra's algorithm to support these features is usually easy.
Extending hierarchical speed-up techniques is often possible, too.
However, the algorithm development is vastly more complex.
For every feature, dedicated research papers exist that extend CH and other approaches.
Supporting the combination of several features is even harder.
For example, we are not aware of any work combining all features mentioned above.
In this article, we describe an algorithmic building block, that allows handling the combination of all above-mentioned features -- and probably more.

@@

There is a lot of work that extends hierarchical speed-up techniques to more complex settings~\cite{bdgmpsww-rptn-16}.
For example, in~\cite{gv-errnt-11} turn information is integrated into CH.
The proposed approach achieves fast queries, but preprocessing becomes an order of magnitude slower compared to classical CH.

@@

In this section, we recall the central aspects of Dijkstra's classical shortest path algorithm~\cite{d-ntpcg-59} and introduce notation used throughout this paper.
Dijkstra's algorithm maintains an array of tentative distances $D$ and a priority queue $Q$ of nodes ordered by increasing distance from $s$.
Initially, all distances are set to $\infty$.
The start node distance $D[s]$ is set to zero and the queue is initialized with $s$.
In each iteration, the closest remaining node $u$ is popped from the queue and \emph{settled}.
Outgoing edges $uv$ are relaxed, i.e. $D[v]$ is set to $\min(D[v], D[u] + w(uv))$.
If the tentative distance at the head node $D[v]$ was improved, it's added to the queue or its position in the queue is adjusted.
Once $t$ was settled, $D[t]$ contains the shortest path distance between $s$ and $t$.
The search can stop when $t$ is settled.

@@

In a CH, each node gets a unique rank representing its ``importance''.
By convention, important nodes are usually depicted at the ``top'' while unimportant nodes are positioned at the ``bottom''.
The intuition is that dead-ends are unimportant while highways or bridges are important.
An edge goes \emph{up} when its tail node has lower rank than its head node.
\emph{Down} edges are defined analogously.
An \emph{up-down path} is a path where only one node $m$ is more important than both its neighbors.
$m$ is called the \emph{mid} node.
An \emph{up path} is a path where the last node is the mid node.
Similarly, the first node is the mid node of a \emph{down path}.
% Every up and down path is an up-down path.
%
In the preprocessing phase, first, a node ordering is computed heuristically.
Then, \emph{shortcut} edges are added to the input graph $G$ to obtain $G^+$.
This is done by repeatedly contracting the least important node and adding shortcuts between its neighbors.
%See~\cite{gssv-erlrn-12} for the details.
After the preprocessing, for every pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path in $G^+$ with the same length as a shortest path in $G$.
See Figure~\ref{fig:ch} for a proof sketch.
From every shortest path (red) in $G$, an up-down path of equal length in $G^+$ (blue) can be constructed.
Thus, we can restrict our search to up-down paths in $G^+$.
The search is bidirectional.
The forward search starts from $s$ and only follows up-edges.
Similarly, the backward search starts at $t$ and only follows down-edges in reversed direction.
The two searches meet at the mid node.
Pseudocode for the backward search, i.e., the path from $m$ to $t$, is presented in Algorithm~\ref{algo:ch-backward}.
The forward search works analogously.
%
A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
On road networks, this is the case~\cite{gssv-erlrn-12,dgpw-crprn-13}.
On graphs with low treewidth, this is also the case~\cite{dsw-cch-15,hs-gbpo-18}.

@@

Some hierarchical techniques were extended to support a three-phase setup.
The phases consist of a slow preprocessing phase, a faster \emph{customization} phase, and fast queries.
The preprocessing phase typically takes hours, the customization seconds, and the query milliseconds.
The preprocessing phase is independent of the weights and only exploits the graph topology.
Weights are introduces in the customization phase.
Paths are computed in the query phase.
This setup is useful to allow for weight changes that are common to all vehicles such as life traffic updates.
The first technique to follow this setup is CRP~\cite{dgpw-crprn-13}.

Customizable Contraction Hierarchies~\cite{dsw-cch-15} (CCH) are a variant of Contraction Hierarchies that uses this three-phase setup.
The key idea is to use a node importance order that is independent of the weight function.
A CCH is also a CH.
The final result of the preprocessing is also an augmented graph $G^+$.
It fulfills all necessary properties for CH query algorithms.
The CH query algorithms can therefore be applied without modifications.
However, the CCH preprocessing grants some additional useful properties.

@@

PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node (or vice versa, the reverse case works analogously).
This is sometimes denoted as a \emph{all-to-one} (or \emph{one-to-all}) problem.
The preprocessing phase remains the same as for CH.
The query phase is split into two steps.
The first step is analogue to the CH query:
From $t$, all reachable nodes via reversed down-edges are explored.
Algorithm~\ref{algo:ch-backward} shows this first step.
For the second step, PHAST organizes nodes into \emph{levels}.
These levels correspond to the importance ordering but allow multiple nodes in the same level.
No edge must connect two nodes within the same level.
PHAST then iterates over all CH levels from top to bottom.
In each iteration, all up-edges starting within the current level are followed in reverse.
After all levels are processed, the shortest distances from all nodes to $t$ were computed.
Pseudocode is provided in Algorithm~\ref{algo:phast}.
PHAST is faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
It can also easily be parallelized.
% However, we will not consider parallelization in this paper.
We refer to~\cite{dgnw-phast-13} for an in-depth experimental performance analysis.
% Using PHAST, we can also compute a tight A* heuristic.
% In the query phase, we first run PHAST to compute the distances from every node to $t$ with respect to $w_\ell$ and store the result in an array $H$.
% Next, we run $A^*$ and implement the heuristic as a lookup in the array $H$.

@@

RPHAST~\cite{delling_et_al:OASIcs:2011:3266}, short for Restricted PHAST, is a PHAST extension for efficiently computing distances from a smaller set of source nodes to one target node (again, the reversed case works analogously), solving the \emph{many-to-one} problem.
Given a set of source nodes $S$, the first step is to copy all up-edges reachable only through up-edges from the source nodes into a new graph.
This step is called \emph{selection}.
Then, in the query step, a target $t$ is given and the PHAST algorithm is applied but the downward sweep (second step of the PHAST algorithm) is performed only on the restricted subgraph.
RPHAST is particularly effective when many targets are queried for the same source set $S$.
If the source set changes often, selection times may become problematic.
% RPHAST can also be applied to the \emph{many-to-many} problem, computing a distance matrix between sets $S$ and $T$ of nodes.
% Implemented with SIMD instructions it is, to the best of our knowledge, the fastest currently known algorithm for this problem setting.

@@

In this section, we discuss another very specific problem setting which naturally arises from A* heuristics.
Here, the target node is known in the selection step, but the source nodes are only queried one after another.
We denote this problem as the \emph{incremental many-to-one} problem.
The first step is the \emph{target selection} where the target node is provided.
Then, an arbitrary number of source nodes $s_i$ are given one after another.
For each source the distance to the target has to be computed before the next source will be provided.

@@

Analogous to A*, our algorithm stores for every node $v$ a tentative distance $D[v]$.
Additionally, it maintains a minimum priority queue.
Diverging from A*, not all nodes need be pushed, but every node has a tentative distance.

@@

Our algorithm differs from A* when removing a node $u$ from the queue.
A* iterates over the outgoing edges $(uv)$ of $u$ and tries to reduce $D[v]$ by relaxing $(uv)$.
If A* succeeds, $v$'s weight in the queue is set to $D[v]+h_t(v)$.
Our algorithm, however, behaves differently, if $d(v)\le 2$.
Our algorithm determines the longest degree two chain of nodes $u, v_1,\ldots, v_k, w$ such that $d(v_i)=2$ and $d(w) > 2$.
If our algorithm succeeds in reducing $D[v_1]$, it does not push $v_1$ into the queue.
Instead, it iteratively tries to reduce all $D[v_i]$.
If it does not reach $w$, then only $D$ is modified, but no queue action is performed.
If $D[w]$ is modified and $d(w)>2$, $w$'s weight in the queue is set to $D[w]+h_t(w)$.

@@

To avoid the downsides of the average potential, we revisit symmetric bidirectional A* and propose a new pruning criterion.
We describe the idea for the forward search.
The pruning rule for the backward search is analogous.
The central idea consists of using information from the backward search to prune edges in the forward search.
We do not use the average heuristic.
Instead of a strong stopping criterion, we use a pruning rule that gets stronger the longer the search runs.
Such a pruning rule will eventually prune all remaining branches and stop the search.
The stopping criteria for each direction remain the same (unidirectional) as before.
However, usually the search stops early because the queues are empty.

@@

Unless stated differently, for all bidirectional A* variants, we always alternate between removing a node from the forward and the backward queues.
In our experiments, we also evaluate expanding the search with the smaller minimum queue key.
This sounds sensible, but our experiments show in Table~\ref{tab:bidir_switching} that it is never beneficial in terms of running time.

@@

A path with nodes $v_1, v_2,\ldots v_k$ has the following \emph{turn-aware weight}: \[
w_\ell(v_1 v_2) + \sum_{i=2}^{k-1}  w_t(v_{i-1},v_i,v_{i+1}) + w_\ell(v_i v_{i+1})
\]
The objective is to find a path between two given edges with minimum turn-aware weight.
The first term $w_\ell(v_1, v_2)$ is the same for all paths, as it only depends on the source edge.
It can thus be ignored during optimization.

@@

If all travel time functions fulfill the \emph{FIFO property}, this problem can be solved using a straightforward extension of Dijkstra's algorithm \cite{d-aassp-69}.
The necessary modification to A* is analogous.
Without the FIFO property the problem becomes NP-hard \cite{or-tnp-89}.
% The FIFO property states that it is never beneficial to wait at a node before entering an edge.
The FIFO property states that it is not possible to arrive earlier by departing later.
Formally stated, the following must hold $\forall e\in E,\tau\in \mathbb{R},\delta\in \mathbb{R}^+: w_q(e, \tau) \le w_q(e, \tau+\delta) + \delta$.
Our implementation stores edge travel times using piece-wise linear functions.
The A* search uses the tentative distance $\tau$ at a node $u$ when evaluating the travel time of outgoing edges $u v$.
This strategy is very similar to TD-ALT~\cite{ndls-bastd-12,dw-lbrdg-07}.

@@

This setup provides an alternative way of integrating live traffic compared to the approach described in Section~\ref{sec:live-predicted-traffic}.
Changes to the live traffic situation are handled by changing $w_\ell$ via a customization.
Additional features such as turn restrictions can be implemented by adjusting $w_q$.
For severe traffic events, the three-phase approach will be significantly faster than the one described in Section~\ref{sec:live-predicted-traffic}.
However, it comes at the expense of a more complex algorithm.
Further, integrating predicted traffic becomes difficult, as $w_q$ may not drop below the $w_\ell$ with integrated congestion.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{fig/lazy_rphast_inc.pdf}
\caption{
Average running times of incremental Lazy RPHAST while querying $|S| = 2^{14}$ from a ball of varying size $|B|$ on OSM Germany excluding selection times.
The upper figure contains the total elapsed running time.
The lower figure contains the averaged running time per source, i.e. $y/x$ from the upper figure.
Note the different y-axis scales and units.
}\label{fig:lazy_rphast_inc}
\Description{
Two line graphs with each eight lines for different ball sizes and algorithms.
The x-axis in both graphs is the number of nodes that were queried up to that point in the incremental query.
In each graph, all lines have a similar shape and appear as scaled variants of each other.
The bigger the ball size the slower the whole algorithm.
The CCH based variant is somewhat slower for the most part.
Only for small ball sizes it is marginally faster when computing distances for more than 4096 sources.
When the ball is the entire graph, the CH based variant takes around 2.5ms for the first 512 sources, 5ms for 2048 sources and 25ms for 16384 sources.
Conversely, the average time per query is around 100mus for the first few queries, then drops to around 5mus after 512 sources, down to 2mus after 16384 sources.
For the smallest ball size of 16384, everything is 1.5 orders of magnitude faster.
}
\end{figure}

@@

Table~\ref{tab:building_blocks} depicts the performance of unidirectional A* with different heuristics and optimizations.
We compare (C)CH-Potentials to three other heuristics.
The first heuristic is, the \emph{Zero} heuristic where $h(v)$ is $0$ for all nodes $v$.
This corresponds to using Dijkstra's algorithm.
Secondly, we compare against our own implementation of ALT~\cite{gw-cppsp-05}.
We use 16 landmarks generated with the avoid strategy~\cite{gw-cppsp-05} and activate all during every query.
Finally, we compare against a hypothetical \emph{Oracle-A*} heuristic.
This heuristic has instant access to a shortest distance array with respect to $w_\ell$, i.e. it is faster than the fastest heuristic possible in our model.
We fill this array before each query using a reverse Dijkstra search from the target node.
In this experiment, we do not account for the running time of Dijkstra's algorithm.
Thus, the reported running times of Oracle-A* do \emph{not} account for any heuristic evaluation.
(C)CH-Potentials compute the same distance estimates, but the heuristic evaluation has some overhead.
Comparing against Oracle-A* allows us to measure this overhead.
No other heuristic, which only has access to the preprocessing weights, can be faster than Oracle-A*.

@@

In Table~\ref{tab:bidir} we investigate the effectiveness of bidirectional search compared to unidirectional search depending on the query weights.
Interestingly, only the zero heuristic and ALT consistently achieve speed-ups through bidirectional search on all query weights.
With $w_q = w_{\ell}$, unidirectional CH-Potentials are already optimal and only traverse the shortest path.
Here, bidirectional search will only introduce unnecessary overhead.
When query weights are scaled up uniformly, bidirectional search achieves some search space reduction, but it is not enough to significantly reduce running times due to the overhead of running a second search.
This changes drastically when the query weight increases are applied non-uniformly in the third scenario.
Here only weights with a speed less than 80\,kph were scaled up.
This touches only the beginning and end of most shortest paths between randomly chosen nodes.
The middle part of the shortest paths will typically use faster edges like highways.
In this case, bidirectional (C)CH-Potentials are a factor of two faster than the unidirectional variant.
The reason for this is that the search space of a unidirectional search expands greatly while exploring the end of the path to the target where the reduced weights are bad.
In contrast, the bidirectional searches meet in the middle of the shortest path where the reduced weights are close to zero and thus avoid this expansion.
This is also the reason why ALT behaves like this for all query weights.
By construction, the ALT heuristic has better reduced weights for edges which are on many shortest paths like highways.
Conversely, unimportant edges have bad reduced weights.
This makes bidirectional search so critical for ALTs performance.
In contrast, having a potential as tight as CH-Potentials makes bidirectional search in many scenarios unnecessary.
Bidirectional search for CH-Potentials only pays of when the reduced weights exhibit certain very specific patterns.

@@

In the other scenarios, the performance of CH-Potentials strongly depends on the quality of the heuristic.
We measure this quality using the length increase of $w_q$ compared to $w_\ell$.
Forbidding highways results in the largest length increase and in the smallest speedup.
The other extreme are turn restrictions.
They have only a small impact on the length increase.
The achieved speedups are therefore comparable to CH speedups.
Since turn costs and restrictions appear mostly in the beginning and end of shortest-paths and not in the middle on highways, utilizing bidirectional search results in even better speed-ups.
Mapbox live traffic has a length increase of around 15\%, which yields running times of 130\,ms.
Applying bidirectional CH-Potentials in this case is actually detrimental to the performance because the bad reduced edge weights appear in the middle of shortest paths.
The same is the case for forbidden tunnels or highways.
Applying CCH-Potentials customized to the current traffic (or the metric with forbidden edges) to these problems yields running times similar to the unmodified case and speed-ups of more than three orders of magnitude.
With bidirectional CCH-Potentials, one can easily additionally support turn costs and still have running times of a few milliseconds without any adjustments to the customization.
